Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-11-03 paper/intro.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@659 3fc257bb-8550-0410-81b3-a5c103da83f7

More recently, MapReduce formulations of traditional number-crunching
kinds of scientific computational tasks have been described, such as
post-processing analysis of simulation data~\cite{Tu08}, graph
algorithmics~\cite{Cohen09}, and linear algebra operations~\cite{Ekanayake09}.
The paper by Tu et al.~\cite{Tu08} was particularly insightful to us,
because it described how MapReduce could be implemented on top of the
ubiquitous distributed-memory message-passing interface (MPI), and how
the intermediate data-shuffle operation is conceptually identical to
the familiar MPI\_Alltoall operation.  Their implementation of
MapReduce was within a Python wrapper to simplify the writing of user
programs.  The paper motivated us to develop our own C++ library built
on top of MPI for use in graph analytics, which we initially released
as open-source software in mid-2009~\cite{MRMPI}.  We have since
worked to optimize several of the library's underlying algorithms and
to enable its operation in out-of-core mode on larger data sets.
These algorithmic improvements are described in this paper and are
part of the current downloadable version~\cite{MRMPI}.

@@

\item {\it In-core or out-of-core operation:} Each MapReduce object
that a processor creates allocates ``pages'' of memory, where the page
size is determined by the user.  Typical MapReduce operations can be
performed using a few such pages (per processor).  If the data set
fits in a single page (per processor), the library performs its
operations in-core.  If the data set exceeds the page size, processors
each write to temporary disk files (on local disks or a parallel file
system) as needed and subsequently read from them.  This out-of-core
mode of operation allows processing of data sets larger than the
aggregate memory of all the processors, i.e. up to the aggregate
available disk space.

@@

\item {\it No fault tolerance or data redundancy:} Current MPI
implementations do not enable easy detection of a dead processor or
retrieval of the data it was working on.  So like most MPI programs, a
parallel program calling the MR-MPI library will hang or crash if a
processor goes away.  Unlike Hadoop, and its HDFS file system which
provides data redundancy, the MR-MPI library reads and writes simple,
flat files.  It can use local per-processor disks, or a parallel file
system, if available, but these typically provide no data redundancy.

>>>>

More recently, MapReduce formulations of traditional number-crunching
kinds of scientific computational tasks have been described, such as
post-processing analysis of simulation data~\cite{Tu08}, graph
algorithmics~\cite{Cohen09}, and linear algebra
operations~\cite{Ekanayake09}.  The paper by Tu et al.~\cite{Tu08} was
particularly insightful to us, because it described how MapReduce
could be implemented on top of the ubiquitous distributed-memory
message-passing interface (MPI), and how the intermediate data-shuffle
operation is conceptually identical to the familiar MPI\_Alltoall
operation.  Their implementation of MapReduce was within a Python
wrapper to simplify the writing of user programs.  The paper motivated
us to develop our own C++ library built on top of MPI for use in graph
analytics, which we initially released as open-source software in
mid-2009.  We have since worked to optimize several of the library's
underlying algorithms and to enable its operation in out-of-core mode
for larger data sets.  These algorithmic improvements are described in
this paper and are part of the current downloadable version; see
Section ~\ref{sec:thanks} for details.

@@

\item {\it In-core or out-of-core operation:} Each MapReduce object
that a processor creates allocates ``pages'' of memory, where the page
size is determined by the user.  MapReduce operations can be performed
using a few such pages (per processor).  If the data set fits in a
single page (per processor), the library performs its operations
in-core.  If the data set exceeds the page size, processors each write
to temporary disk files (on local disks or a parallel file system) as
needed and subsequently read from them.  This out-of-core mode of
operation allows processing of data sets larger than the aggregate
memory of all the processors, i.e. up to the aggregate available disk
space.

@@

\item {\it No fault tolerance or data redundancy:} Current MPI
implementations do not enable easy detection of a dead processor or
retrieval of the data it was working on.  So like most MPI programs, a
parallel program calling the MR-MPI library will hang or crash if a
processor goes away.  Unlike Hadoop, and its file system (HDFS) which
provides data redundancy, the MR-MPI library reads and writes simple,
flat files.  It can use local per-processor disks, or a parallel file
system, if available, but these typically provide no data redundancy.

### 2010-11-03 paper/mr.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@659 3fc257bb-8550-0410-81b3-a5c103da83f7

Note that the first portion of the {\it collate()} operation involves
all-to-all communication (each processor sends and receives data from
every other processor) using a distributed hash table.  The
communication can either be done via a MPI\_Alltoall() library call,
or by a custom routine that aggregates messages and invokes
point-to-point MPI\_Send() and MPI\_IRecv() calls.

@@

Several related MapReduce operations are provided by the library.  For
example, the {\it collate()} function described above calls two other
functions: {\it aggregate()}, which performs the all-to-all
communication, and {\it convert()}, which turns a list of KV pairs
into KMV pairs.  Both functions can be called directly.  The {\it
compress()} function allows on-processor operations to be performed;
it is equivalent to a {\it convert()} with on-processor KVs as input,
followed by a {\it reduce()}.  The {\it clone()} function turns a list
of KV pairs into KMV pairs, with one value per key.  The {\it
collapse()} function turns $N$ KV pairs into one KMV pair, with the
keys and values of the KV pairs becoming $2N$ values of a single
multivalue assigned to a new key.  The {\it gather()} function
collects KV pairs from all processors to a subset of processors; it is
useful for doing output from one or a few processors.  Library calls
for sorting datums by key or value or for sorting the values within
each multivalue are also provided.  These routines invoke the
C-library {\it quicksort()} function to compute the sorted ordering of
the KV pairs (or values within a multivalue), using a user-provided
comparison function.  The KV pairs (or values in a multivalue) are
then copied into a new data structure in sorted order.

The interface to various low-level operations is provided so that a
user's program can string them together in various ways to construct
interesting MapReduce algorithms.  For example, output from a {\it
reduce()} can serve as input to a subsequent {\it map()} or {\it
collate()}.  Likewise, data can be partitioned across multiple
MapReduce (MR) objects, each with their own set of KV pairs.  KV pairs
from multiple MR objects can be combined to perform new sequences of
{\it map()}, {\it collate()}, and {\it reduce()} operations.

The above discussion assumed that the KV or KMV pairs stored by a
processor fit in its physical memory.  In this case, MR-MPI performs
only ``in-core'' processing and no disk files are written or read by
any of the processors, aside from initial input data if it exists or
final output data if it is generated.  Note that the aggregate
physical memory of large parallel machines can be multiple terabytes,
which allows for large data sets to be processed in-core, assuming the
KV and KMV pairs remain evenly distributed across processors
throughout the sequence of MapReduce operations.  The use of hashing
to assign keys to processors typically provides for such
load-balancing.  In the next section, we discuss what happens when
data sets do not fit in available memory.

>>>>

Note that the first portion of the {\it collate()} operation involves
all-to-all communication since each processor exchanges KV pairs with
every other processor.  The communication can either be done via a
MPI\_Alltoall() library call, or by a custom routine that aggregates
datums into large messages and invokes point-to-point MPI\_Send() and
MPI\_IRecv() calls.

@@

Several related MapReduce operations are provided by the library.  For
example, the {\it collate()} function described above calls two other
functions: {\it aggregate()}, which performs the all-to-all
communication, and {\it convert()}, which reorganizes a set of KV
pairs into KMV pairs.  Both functions can be called directly.  The
{\it compress()} function allows on-processor operations to be
performed; it is equivalent to a {\it convert()} with on-processor KVs
as input, followed by a {\it reduce()}.  The {\it clone()} function
turns a list of KV pairs into KMV pairs, with one value per key.  The
{\it collapse()} function turns $N$ KV pairs into one KMV pair, with
the keys and values of the KV pairs becoming $2N$ values of a single
multivalue assigned to a new key.  The {\it gather()} function
collects KV pairs from all processors to a subset of processors; it is
useful for doing output from one or a few processors.  Library calls
for sorting datums by key or value or for sorting the values within
each multivalue are also provided.  These routines invoke the
C-library {\it quicksort()} function to compute the sorted ordering of
the KV pairs (or values within a multivalue), using a user-provided
comparison function.  The KV pairs (or values in a multivalue) are
then copied into a new data structure in sorted order.

The interface to various low-level operations is provided so that a
user's program can string them together in various ways to construct
usefule MapReduce algorithms.  For example, output from a {\it
reduce()} can serve as input to a subsequent {\it map()} or {\it
collate()}.  Likewise, data can be partitioned across multiple
MapReduce (MR) objects, each with their own set of KV pairs.  For
example, as will be illustrated in Section ~\ref{sec:graph}, one MR
object can store the edges of a graph, and another its vertices.  KV
pairs from multiple MR objects can be combined to perform new
sequences of {\it map()}, {\it collate()}, and {\it reduce()}
operations.

The above discussion assumed that the KV or KMV pairs stored by a
processor fit in its physical memory.  In this case, the MR-MPI
library performs only ``in-core'' processing and no disk files are
written or read by any of the processors, aside from initial input
data (if it exists) or final output data (if it is generated).  Note
that the aggregate physical memory of large parallel machines can be
multiple terabytes, which allows for large data sets to be processed
in-core, assuming the KV and KMV pairs remain evenly distributed
across processors throughout the sequence of MapReduce operations.
The use of hashing to assign keys to processors typically provides for
such load-balancing.  In the next section, we discuss what happens
when data sets do not fit in available memory.

### 2010-11-03 paper/outcore.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@659 3fc257bb-8550-0410-81b3-a5c103da83f7

We now explain how the MapReduce operations described in the previous
section work in out-of-core mode.  The {\it map()} and {\it reduce()}
operations are relatively simple.  As a {\it map()} generates KV pairs
via the user function, a page of memory fills up, one KV pair at a
time.  When the page is full, it is written to disk.  If the source of
data for the {\it map()} operation is an existing set of KV pairs,
those datums are read, one page at a time, and a pointer to each KV
pair is given to the user function.  Similarly, a {\it reduce()} reads
one page of KMV pairs from disk, and passes a pointer to each pair,
one at a time, to the user function that typically generates new KV
pairs.  The generated pairs fill up a new page, which is written to
disk when full, just as with the {\it map()} operation.  Thus for both
a {\it map()} and {\it reduce()}, out-of-core disk files are read and
written sequentially, one page at a time, which requires at most two
pages of memory.

@@

Performing an our-of-core {\it collate()} operation is more complex.
Recall that the operation occurs in two stages.  First, keys are
hashed to ``owning'' processors and the KV datums are communicated to
new processors in an all-to-all fashion.  In out-of-core mode, this
operation is performed on one page of KV data at a time.  Each
processor reads in a page of KV pairs, the communication pattern
(which processors receive which datums in that page) is determined,
and all-to-all communication is performed.  Each processor allocates a
two-page chunk of memory to receive incoming KV pairs.  On average
each processor should receive one page of KV pairs; the two-page
allocation allows for some load imbalance.  If the KV pairs are
distributed unevenly so that this limit is exceeded, the all-to-all
communication is performed in smaller, multiple passes until the full
page contributed by every processor has been communicated.  Performing
an MPI\_Alltoall(), or using the custom all-to-all routines provided
by the MR-MPI library, requires allocation of auxiliary arrays that
store processor indices, datum pointers, and datum lengths.  For the
case of many tiny KV datums, the total number of memory pages required
to perform the all-to-all communication, including the source and
target KV pages is at most seven.

@@

Other MR-MPI library calls discussed in the previous section, such as
{\it clone()}, {\it collapse()}, and {\it gather()}, can also operate
in out-of-core mode, typically with one pass through the KV or KMV
data and the use of one or two memory pages.  Sorting KV pairs, by key
or value, is an exception.  An out-of-core merge sort is performed as
follows.  Two pages of KV datums are read from disk.  Each is sorted
using the C-library quicksort() function, as discussed in the previous
section.  The two pages are then scanned and merged into a new file
which is written to disk as its associated memory page fills up.  This
process requires five memory pages, which includes vectors of KV datum
pointers and lengths needed by the in-memory quicksort() operation.
Once all pairs of pages have been merged, the sort continues in a
recursive fashion, merging pairs of files into a new third file,
without the need to perform additional in-memory sorts.  Thus the
overall memory requirement is five pages.  The number of read/write
passes through the KV data set for the merge sort is $O(\log_2{M})$,
where $M$ is the number of pages of KV pairs.  The number of passes
could be reduced at the cost of allocating more in-memory pages.

>>>>

We now explain how the MapReduce operations described in the previous
section work in out-of-core mode.  The {\it map()} and {\it reduce()}
operations are relatively simple.  As a {\it map()} generates KV pairs
via the user callback function, a page of memory fills up, one KV pair
at a time.  When the page is full, it is written to disk.  If the
source of data for the {\it map()} operation is an existing set of KV
pairs, those datums are read, one page at a time, and a pointer to
each KV pair is given to the user function.  Similarly, a {\it
reduce()} reads one page of KMV pairs from disk, and passes a pointer
to each pair, one at a time, to the user callback function that
typically generates new KV pairs.  The generated pairs fill up a new
page, which is written to disk when full, just as with the {\it map()}
operation.  Thus for both a {\it map()} and {\it reduce()},
out-of-core disk files are read and written sequentially, one page at
a time, which requires at most two pages of memory.

@@

Performing an our-of-core {\it collate()} operation is more complex.
Recall that the operation occurs in two stages.  First, keys are
hashed to ``owning'' processors and the KV pairs are communicated to
new processors in an all-to-all fashion.  In out-of-core mode, this
operation is performed on one page of KV data (per processor) at a
time.  Each processor reads in a page of KV pairs, the communication
pattern is determined (which processors receive which datums in that
page), and all-to-all communication is performed.  Each processor
allocates a two-page chunk of memory to receive incoming KV pairs.  On
average each processor should receive one page of KV pairs; the
two-page allocation allows for some load imbalance.  If the KV pairs
are distributed unevenly so that this limit is exceeded, the
all-to-all communication is performed in smaller, multiple passes
until the full page contributed by every processor has been
communicated.  Performing an MPI\_Alltoall(), or using the custom
all-to-all routines provided by the MR-MPI library, requires
allocation of auxiliary arrays that store processor indices, datum
pointers, and datum lengths.  For the case of many tiny KV pairs, the
total number of memory pages required to perform the all-to-all
communication, including the source and target KV pages is at most
seven.

@@

Other MR-MPI library calls discussed in the previous section, such as
{\it clone()}, {\it collapse()}, and {\it gather()}, can also operate
in out-of-core mode, typically with one pass through the KV or KMV
data and the use of one or two memory pages.  Sorting KV pairs, by key
or value, is an exception.  An out-of-core merge sort is performed as
follows.  Two pages of KV datums are read from disk.  Each is sorted
in local memory using the C-library quicksort() function, as discussed
in the previous section.  The two pages are then scanned and merged
into a new file which is written to disk as its associated memory page
fills up.  This process requires five memory pages, which includes
vectors of KV datum pointers and lengths needed by the in-memory
quicksort() operation.  Once all pairs of pages have been merged, the
sort continues in a recursive fashion, merging pairs of files into a
new third file, without the need to perform additional in-memory
sorts.  Thus the overall memory requirement is five pages.  The number
of read/write passes through the KV data set for the merge sort is
$O(\log_2{M})$, where $M$ is the number of pages of KV pairs.  The
number of passes could be reduced at the cost of allocating more
in-memory pages.
